---
title: "Kruskal-Wallis is (almost) a one-way ANOVA on ranked data"
author: "Jonas Kristoffer Lindeløv"
date: "September 5, 2018"
output: 
  html_document:
    number_sections: yes
    self_contained: no
    toc: yes
---

<!-- from https://stackoverflow.com/a/37839683/1297830 -->
<link rel="stylesheet" type="text/css" href="hideOutput.css">
<script src="hideOutput.js"></script>

This document presents the close relationship between the Kruskal-Wallis test and one-way ANOVAs. Namely, that Kruskal-Wallis test is, to a close approximation, just a one-way ANOVA on $rank$ed $y$. It is an appendix to the post "[Common statistical tests as linear models](https://rpubs.com/lindeloev/test_as_linear)".

**TL;DR: Below, I argue that this approximation is good enough when the sample size is 12 or greater and virtually perfect when the sample size is 30 or greater.**


# Simulate data
Calculate statistics for different sample sizes ($N$) and differences between groups ($mu$).

<div class="fold s o">
```{r simulate, message=FALSE, warning=FALSE}
library(tidyverse)

# Parameters
Ns = c(seq(from=6, to=20, by=2), 30, 50, 80)
mus = c(0, 0.5, 1)  # Means
PERMUTATIONS = 1:200

# Run it
D = expand.grid(set=PERMUTATIONS, mu=mus, N=Ns) %>%
  mutate(
    # Generate data. One normal and one weird
    value = map2(N, mu, ~c(rnorm(.x), rnorm(.x) + .y, rnorm(.x))),
    group = map(N, ~factor(c(rep('A', .x), rep('B', .x), rep('C', .x)))),
    
    # Tests
    kruskal_raw = map2(value, group, ~ kruskal.test(.x ~ .y)),
    anova_raw = map2(value, group, ~aov(rank(.x) ~ .y)),
    
    # Tidy it up
    kruskal = map(kruskal_raw, broom::tidy),
    anova = map(anova_raw, broom::glance)
  ) %>%
  
  # Get as columns instead of lists; then remove "old" columns
  unnest(kruskal, anova, .sep='_') %>%
  select(-value, -group, -kruskal_raw, -anova_raw)

head(D)
```
</div>

# Show results
I have found that the effect size ($mu$) makes no difference, so I collapse the analyses below across $mu$.

<div class="fold s">
```{r, fig.width=10, fig.height=3}
D$N = factor(D$N)  # Make N a factor for prettier plotting

library(ggplot2)
library(patchwork)

# A straight-up comparison of the p-values
p_relative = ggplot(D, aes(x=kruskal_p.value, y=anova_p.value, color=N)) + 
  geom_line() + 
  geom_vline(xintercept=0.05, lty=2) +
  geom_hline(yintercept=0.05, lty=2) +
  
  labs(title='Absolute relation', x = 'Kruskal-Wallis p-value', y = 'ANOVA p-value') + 
  #coord_cartesian(xlim=c(0, 0.10), ylim=c(0, 0.11)) + 
  theme_gray(13) + 
  guides(color=FALSE)

# Looking at the difference (error) between p-values
p_error_all = ggplot(D, aes(x=kruskal_p.value, y=anova_p.value-kruskal_p.value, color=N)) + 
  geom_line() + 
  geom_vline(xintercept=0.05, lty=2) +
  
  labs(title='Error', x = 'Kruskal-Wallis p-value', y = 'T-test p-value deviation') + 
  theme_gray(13) + 
  guides(color=FALSE)

# Same, but zoomed in around p=0.05
p_error_zoom = ggplot(D, aes(x=kruskal_p.value, y=anova_p.value-kruskal_p.value, color=N)) + 
  geom_line() + 
  geom_vline(xintercept=0.05, lty=2) +
  
  labs(title='Zoomed error', x = 'Kruskal-Wallis p-value', y = 'T-test p-value deviation') + 
  coord_cartesian(xlim=c(0, 0.10), ylim=c(-0.020, 0.000)) + 
  theme_gray(13)

# Show it. Patchwork is your friend!
p_relative + p_error_all + p_error_zoom
```
</div>


# Conclusion
Reasonably accurate when N = 12 or greater.
Almost exact when N = 30.